\textbf{The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models}

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/02-Challenges.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
# linear response, correlated errors (subject effect)
par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

>>>>

```{r, l-q-response, fig.width=10, fig.height=10, out.width='75%',fig.align='center', echo=FALSE,message=FALSE,fig.show='hold',fig.cap ='(ref:l-q-response-caption)'}
# linear response, correlated errors (subject effect)
#par(mar = c(0.5, 0.5, 0.5, 0.5))
A/C+plot_annotation(tag_levels = 'A')
```

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/04-Longitudinal analysis with GAMs.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#E1DD37FF') +
    geom_line(color='#E75B64FF',size=1) +
    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
    geom_ribbon(data=comp_StO2_full%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
    geom_ribbon(data=comp_StO2_full %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#E1DD37FF') +
    geom_line(color='#E75B64FF',size=1) +
    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
    geom_ribbon(data=comp_StO2_missing%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
    geom_ribbon(data=comp_StO2_missing %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

## Determination of significance in GAMs for longitudinal data{#GAM-sig}

>>>>

c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#DB3A07FF') +
    geom_line(color='#E75B64FF',size=1) +
    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
    geom_ribbon(data=comp_StO2_full%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
    geom_ribbon(data=comp_StO2_full %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#DB3A07FF') +
    geom_line(color='#E75B64FF',size=1) +
    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
    geom_ribbon(data=comp_StO2_missing%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
    geom_ribbon(data=comp_StO2_missing %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

## Determination of significance in GAMs for longitudinal data{#GAM-significance}

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/06-Appendix.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

# (APPENDIX) Appendix {-}

@@

@@

```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}
#need to add figure number and caption
appr1<-appraise(gam_00)
sm1<-draw(gam_00)
visual_check<-appr1+sm1

visual_check 
```

@@

visual_check2
```

@@

visual_check3
```

@@

The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance). In this case, the "design matrix" is used to estimate the pairwise comparisons (see Manuscript for details and references that discuss this). Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group. This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).

@@

#plot the difference
c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
  #ribbon for difference confidence interval  
  geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='black') +
    geom_line(color='black',size=1) +
    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
  #highlight area under the curve where "Control" is higher
  geom_ribbon(data=comp_StO2%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='blue') +
  #highlight area under the curve where "Treatment" is higher
  geom_ribbon(data=comp_StO2 %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='red') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

(ref:pairwise-comp-caption) Smooth pairwise comparisons using a 95% confidence interval for the difference between smooths.

```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pariwise-comp-caption)'}
c1
```

Of notice, a conveniente wrapper for the function described above exists as `difference_smooths` in the package `gratia`. It produces the same plot when the function is used on a fitted model. The function syntax can be checked at:

@@

This code chunk randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted. Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.

@@

```{r,pairwise-comp-Appendix,echo=FALSE,include=FALSE,message=FALSE,warning=FALSE}
##Pairwise comparisons

>>>>

---
output:
  pdf_document: default
  html_document: default
---
# (APPENDIX) Appendix {-}

@@

```{r, plotting layouts,echo=TRUE, include=TRUE}
layout1 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2)
)

layout2 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2),
  area(1,4,2),
  area(1,5,1)
)

@@

```{r, first-GAM-diag, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:first-GAM-caption)'}

appr1<-appraise(gam_00)
sm1<-draw(gam_00)
visual_check<-appr1+sm1

visual_check+plot_layout(design=layout1)
```

@@

visual_check2+plot_layout(design=layout1)
```

@@

visual_check3+plot_layout(design=layout2)
```

@@

The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \@ref(GAM-significance). In this case, the "design matrix" is used to estimate the pairwise comparisons (see main manuscript for details and associated references). Briefly, the "design matrix" (also known as the "Xp matrix") from the selected model (`gam1`) is used to calculate a 95% confidence interval of the difference between the smooth terms for each group. This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).

@@

#plot the difference
c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
  #ribbon for difference confidence interval  
  geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#DB3A07FF') +
    geom_line(color='black',size=1) +
    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
  #highlight area under the curve where "Control" is higher
  geom_ribbon(data=comp_StO2%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
  #highlight area under the curve where "Treatment" is higher
  geom_ribbon(data=comp_StO2 %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

@@

(ref:pairwise-comp-caption) Smooth pairwise comparisons for model `gam1` using a 95% confidence interval for the difference between smooths.

```{r, pairwise-comp-workflow-fig, fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:pairwise-comp-caption)'}
c1
```

Of notice, a convenient wrapper for the function described above exists in the package `gratia`. In this package, `difference_smooths` is a function that makes the comparisons and produces Figure \@ref(fig:pairwise-comp-workflow-fig) when is used on a fitted model. The function syntax and an example can be found at:

@@

This code chunk first randomly deletes 40% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data. Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.

@@

```{r,pairwise-comp-Appendix,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}
##Pairwise comparisons

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.Rmd
Need to improve aesthetics of combined gam plots in the Appendix

#global themes for plotting
thm<-scale_color_viridis_d(option="magma",begin=0.1,end=0.7)
thm1<-scale_fill_viridis_d(option="magma",begin=0.1,end=0.7)#global theme for plotting

>>>>

#global themes for plotting
thm<-scale_color_viridis_d(option="turbo",begin=0,end=1)
thm1<-scale_fill_viridis_d(option="turbo",begin=0,end=1)#global theme for plotting

### 2021-05-24 Manuscripts/Manuscript by chapters -v1/Full_document.tex
Need to improve aesthetics of combined gam plots in the Appendix

\hypertarget{determination-of-significance-in-gams-for-longitudinal-data}{%
\subsection{Determination of significance in GAMs for longitudinal data}\label{determination-of-significance-in-gams-for-longitudinal-data}}

@@

@@

\begin{lstlisting}
## Error: The given dimensions cannot hold all panels. Please increase `ncol` or `nrow`
\end{lstlisting}

@@

@@

This code chunk randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted. Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.

@@

>>>>

\hypertarget{GAM-significance}{%
\subsection{Determination of significance in GAMs for longitudinal data}\label{GAM-significance}}

@@

\begin{lstlisting}[language=R]
layout1 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2)
)

layout2 <- c(
  area(1, 1),
  area( 1, 2),
  area(2, 1),
  area(2, 2),
  area(1, 3, 2),
  area(1,4,2),
  area(1,5,1)
)
\end{lstlisting}

@@

@@

\hypertarget{pairwise-comparisons-of-smooth-confidence-intervals}{%
\paragraph{Pairwise comparisons of smooth confidence intervals}\label{pairwise-comparisons-of-smooth-confidence-intervals}}

The estimation of significant differences between each treatment group can be achieved via pairwise comparisons of the smooth confidence intervals as described in section \ref{GAM-significance}. In this case, the ``design matrix'' is used to estimate the pairwise comparisons (see main manuscript for details and associated references). Briefly, the ``design matrix'' (also known as the ``Xp matrix'') from the selected model (\passthrough{\lstinline!gam1!}) is used to calculate a 95\% confidence interval of the difference between the smooth terms for each group. This approach allows to estimate the time intervals where a significant difference exists between the groups (confidence interval above or below 0).

\begin{lstlisting}[language=R]
##Pairwise comparisons

##matrix that contains the basis functions evaluated at the points in pdat
    xp <- predict(gam1, newdata = pdat, type = 'lpmatrix')

#Find columns in xp where the name contains "Control"
    c1 <- grepl('Control', colnames(xp))

#Find columns in xp where the name contains 'Treatment'
    c2 <- grepl('Treatment', colnames(xp))

#Find rows in pdat that correspond to either 'Control' or 'Treatment'
    r1 <- with(pdat, Group == 'Control')
    r2 <- with(pdat, Group == 'Treatment')

# In xp: find the rows that correspond to Control or Treatment, those that do not match will be
    #set to zero. Then, substract the values from the rows corresponding to 'Control' from those that correspond
    #to 'Treatment'
    X <- xp[r1, ] - xp[r2, ]

    ## remove columns that do not contain name 'Control' or 'Treatment'
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols, those that do not contain in the characters 's('
    X[, !grepl('^s\\(', colnames(xp))] <- 0

    #Multiply matrix by model coefficients. X has (p,n) (rows, columns) and the coefficient matrix has
    #dimensions (n,1). The resulting matrix has dimensions (p,1)
    dif <- X %*% coef(gam1)

    #comp<-test %*% coef(gam1)[3:10]

#Calculate standard error for the computed differences using the variance-covariance matrix
    #of the model
    se <- sqrt(rowSums((X %*% vcov(gam1, unconditional = FALSE)) * X))
    crit <- qt(0.05/2, df.residual(gam1), lower.tail = FALSE)
    #upper  limits
    upr <- dif + (crit * se)
    #lower limits
    lwr <- dif - (crit * se)
    #put all components in a dataframe for plotting
    comp1<-data.frame(pair = paste('Control', 'Treatment', sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)

#add time point sequence
comp_StO2 <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp1))

#plot the difference
c1<-ggplot(comp_StO2, aes(x = Day, y = diff, group = pair)) +
  #ribbon for difference confidence interval  
  geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#DB3A07FF') +
    geom_line(color='black',size=1) +
    geom_line(data=comp_StO2,aes(y=0),size=0.5)+
  #highlight area under the curve where "Control" is higher
  geom_ribbon(data=comp_StO2%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
  #highlight area under the curve where "Treatment" is higher
  geom_ribbon(data=comp_StO2 %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )
\end{lstlisting}

{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/pairwise-comp-workflow-fig-1}

\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:pairwise-comp-workflow-fig}
\end{figure}

Of notice, a convenient wrapper for the function described above exists in the package \passthrough{\lstinline!gratia!}. In this package, \passthrough{\lstinline!difference\_smooths!} is a function that makes the comparisons and produces Figure \ref{fig:pairwise-comp-workflow-fig} when is used on a fitted model. The function syntax and an example can be found at:

\url{https://cran.r-project.org/web/packages/gratia/gratia.pdf}

@@

This code chunk first randomly deletes 40\% of the total observations in the original simulated data, and then an interaction GAM is fitted to the remaining data. Model diagnostics are presented, and an object that stores the fitted smooths is saved to be called in the final code chunk to build the figure.

@@

\hypertarget{pairwise-comparisons-in-gams-full-and-missing-data-cases}{%
\subsection{Pairwise comparisons in GAMs: full and missing data cases}\label{pairwise-comparisons-in-gams-full-and-missing-data-cases}}

The next code chunk reproduces Figure \ref{fig:plot-pairwise-comp}. Here pairwise comparisons are made for the full and missing datasets.

\begin{lstlisting}[language=R]
##Pairwise comparisons

pdat <- expand.grid(Day = seq(0, 10, length = 400),
                    Group = c('Control', 'Treatment'))

#this function takes the model, grid and groups to be compared using the lpmatrix

smooth_diff <- function(model, newdata, g1, g2, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    #Find columns in xp where the name contains "Control"
    col1 <- grepl(g1, colnames(xp))
    #Find columns in xp where the name contains 'Treatment'
    col2 <- grepl(g2, colnames(xp))
    #r1 <- newdata[[var]] == f1
    #r2 <- newdata[[var]] == f2
    row1 <- with(newdata, Group == g1)
    row2 <- with(newdata, Group == g2)
    ## difference rows of xp for data from comparison
    X <- xp[row1, ] - xp[row2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (col1 | col2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(g1, g2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}

comp1<-smooth_diff(gam1,pdat,'Control','Treatment')

comp_StO2_full <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp1)) %>%
  mutate(interval=case_when(
    upper>0 & lower<0~"no-diff",
    upper<0~"less",
    lower>0~"greater"
  ))

c1<-ggplot(comp_StO2_full, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#DB3A07FF') +
    geom_line(color='#E75B64FF',size=1) +
    geom_line(data=comp_StO2_full,aes(y=0),size=0.5)+
    geom_ribbon(data=comp_StO2_full%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
    geom_ribbon(data=comp_StO2_full %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

###for missing data
comp2<-smooth_diff(mod_m1,pdat,'Control','Treatment')
comp_StO2_missing <- cbind(Day = seq(0, 10, length = 400),
                   rbind(comp2))

missing_plot<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line(color='black',size=1) +
    facet_wrap(~ pair) +
    labs(x = 'Days', 
         y = expression(paste('Difference in StO'[2],'\n (missing data)' 
                              )))+
  scale_x_continuous(breaks=c(0,2,5,7,10))+
  theme_classic()+
  theme(
     text=element_text(size=18),
     legend.title=element_blank()
     )

c2<-ggplot(comp_StO2_missing, aes(x = Day, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper),
                alpha = 0.5,
                fill='#DB3A07FF') +
    geom_line(color='#E75B64FF',size=1) +
    geom_line(data=comp_StO2_missing,aes(y=0),size=0.5)+
    geom_ribbon(data=comp_StO2_missing%>%
                    filter(lower>0),
                aes(ymin =0, ymax =lower),
                alpha = 0.5,
                fill='#30123BFF') +
    geom_ribbon(data=comp_StO2_missing %>%
                    filter(upper<0),
                    aes(ymin =0, ymax =upper),
                alpha = 0.5,
                fill='#7A0403FF') +
    facet_wrap(~ pair) +
    theme_classic()+
    labs(x = 'Days', y = expression(paste('Difference in StO'[2] )))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
    theme(
        text=element_text(size=18),
        legend.title=element_blank()
    )

pair_comp<-c1+c2
\end{lstlisting}

Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths. Pairwise comparisons for smooth terms. A: Pairwise comparisons for the full dataset. B: Pairwise comparisons for the dataset with missing observations. Significant differences exist where the interval does not cover 0. In both cases the effect of treatment is significant after day 5.

{\centering \includegraphics[width=0.75\linewidth,]{Full_document_files/figure-latex/plot-pairwise-comp-Appendix-1}

\caption{Smooth pairwise comparisons for model \passthrough{\lstinline!gam1!} using a 95\% confidence interval for the difference between smooths.}\label{fig:plot-pairwise-comp-Appendix}
\end{figure}
